---
title: "JCR P-Values"
description: |
  A short exercise in extracting p values from academic papers.
author:
  - name: Ethan Milne
    url: https://approachingsignificance.com
date: 2022-03-03
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
```

# Required Packages

```{r message = FALSE, warning = FALSE}
library(pdftools)
library(tidyverse)
```

# Extract P Values

The code chunk below extracts the P values from papers in the following way:

* Get the names of all papers in a folder containing them
* Loop through each paper name and use the `pdftools` package to extract their raw text
* Do some basic cleaning (e.g. getting rid of "\\n", which denote paragraph breaks)
* Extract all strings of text that are preceded by `"P >"`, `"P <"`, `"P ="`, and which end in a closed bracket `")"`
* Output a data frame with 2 columns: `P values` and `Paper Name`

```{r}
files <- list.files("PDFs")

Results <- data.frame(matrix(ncol = 2, nrow = 0))
names(Results) <- c("P_Value", "Paper")

for (i in 1:length(files)) {
  name <- files[i]
  
  text <-
    pdf_text(
        here::here("_posts", "2022-03-03-jcr-p-values", "PDFs", name)
    )
  text <- gsub("\n", "", text)
  text <- gsub("  ", "", text)
  
  values <-
    unlist(str_extract_all(text, 'p\\s?[=<>]\\s?\\.\\d{1,4}')) |>
    as.data.frame() |>
    mutate(Paper = name)
  
  if (is_empty(values)) {
    next
  } else{
    names(values) <- c("P_Value", "Paper")
    
    Results <- rbind(values, Results)
  }
}
```

# Cleaning

Now that I have every P value, I need to extract the actual number. This is moderately challenging - P values are often reported without the leading `0` (e.g. p = .07), and a p value that is reported as **greater than** 0.05 is different from one that is **equal** to or **less than**, and those differences need to be recorded somewhere for any future work I may do.

In summary, what the below code does is:

* Extract the "raw numeric value" from each reported p value string
* Replace the prior "p [<=>]" with a "0" instead
* Convert this column to numeric

```{r}
Cleaned_Results <- Results |>
  mutate(
    Raw_Value = P_Value,
    Raw_Value = gsub("p < ", "0", Raw_Value),
    Raw_Value = gsub("p = ", "0", Raw_Value),
    Raw_Value = gsub("p > ", "0", Raw_Value),
    Raw_Value = gsub("p >", "0", Raw_Value),
    Raw_Value = gsub("p =", "0", Raw_Value),
    Raw_Value = gsub("p <", "0", Raw_Value),
    Raw_Value = gsub("p< ", "0", Raw_Value),
    Raw_Value = gsub("p= ", "0", Raw_Value),
    Raw_Value = gsub("p> ", "0", Raw_Value),
    Raw_Value = gsub("p<", "0", Raw_Value),
    Raw_Value = gsub("p=", "0", Raw_Value),
    Raw_Value = gsub("p>", "0", Raw_Value)
  ) |>
  mutate(Operator = str_extract(P_Value, "[=<>]"))

Cleaned_Results$Raw_Value <- as.numeric(Cleaned_Results$Raw_Value)
```



# Plotting

Finally, I wanted to plot the raw p values I've found. There's little analytic code here - mostly just ggplot aesthetic wrangling. I used the `{ggthemes}` package for R to get the `theme_fivethirtyeight` function which gives me a lot of aesthetic power, for lack of a better term. 

I've added X axis breaks at the standard p value thresholds - 0.1, 0.05, 0.01, 0.001. We should expect that there is significant clustering around these thresholds, as most researchers seem to report inequalities (p < x) rather than exact values.

Finally, p values aren't exactly linear in their distribution, so everything is put on a log scale for easier interpretation.

```{r warning = FALSE, message = FALSE}
library(ggthemes)
library(ggtext)
library(ggpubr)

ggplot(Cleaned_Results, aes(x = Raw_Value)) +
  geom_histogram(bins = 50, fill = "black") +
  scale_x_log10(breaks = c(0.0001, 0.001, 0.01, 0.05, 0.1, 1),
                labels = c(".0001", ".001", ".01", ".05", ".1", "1")) +
  theme_pubclean() +
  theme(
    axis.title.x = element_text(),
    axis.title.y = element_text(),
    plot.title = element_markdown(),
    plot.background = element_rect(colour = "black", fill=NA, size=2)
  ) +
  labs(
    title = "<span style = 'color: #ed713a;'>Distribution of P Values</span>",
    subtitle = "Appearing in JCR, 2019-2020 editions",
    x = "P Values",
    y = "Count"
  )
```
